### 2015-12-01 Manuscript/BiologyLetters/Missing_mammals-draft.tex
Final minor typos edits
docx conversion

There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{eernissetaxonomic1993,ronquista2012}, combines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), producing a phylogeny with living and fossil taxa at the tips. 
A downside of this method is that it requires molecular data for living taxa and morphological data for both living and fossil taxa.
Chunks of this data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may restrict the amount of morphological data available.
Additionally, it has become less common to collect morphological characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded morphological data).
Unfortunately this missing data can lead to errors in phylogenetic inference.
Simulations show that the ability of the Total Evidence method to recover the correct topology decreases when there is little overlap between morphological data in living and fossil taxa, and that the effect of missing data on topology is greatest when living taxa have few morphological data \cite{GuillermeCooper}.
This is because (1) fossils cannot branch in the correct clade if it contains no morphological data for living taxa; and (2) fossils have a higher probability of branching within clades with more morphological data for living taxa, regardless of whether this is the correct clade \cite{GuillermeCooper}.

@@

%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all cladistic matrices containing any living and/or fossil mammal taxa from three major public databases: MorphoBank (\url{http://www.morphobank.org/} \cite{morphobank}), Graeme Lloyd's website (\url{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\url{https://github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search for matrices that were not uploaded to these databases (see Supplementary Materials Section 1 for a detailed description of the search procedure).
In total, we downloaded 286 matrices containing 5228 unique operational taxonomic units (OTUs). 
We used OTUs rather than species since entries in the matrices ranged from species to families, and standardised the taxonomy as described in Supplementary Materials (section 1). 
We designated as ``living'' all OTUs that were either present in the phylogeny of \cite{BinindaEmonds} or the taxonomy of \cite{wilson2005mammal}.

Matrices with few characters are problematic when comparing available data among matrices because (1) they have less chance of having characters that overlap with those of other matrices \cite{wagner2000} and (2) they are more likely to contain a higher proportion of specific characters that are not-applicable across large clades (e.g. ``antler ramifications'' is a character that is only applicable to Cervidae not all mammals \cite{Brazeau2011}).
Therefore we selected only matrices containing \textgreater 100 characters for each OTU.
This threshold was chosen to correspond with the number of characters used in \cite{GuillermeCooper} and \cite{harrisonamong-character2014}.
Results of analyses with no threshold are available in Supplementary Material. 
After removing matrices with \textless 100 characters, we retained 1074 unique living mammal OTUs from 126 matrices. % 1601 unique living OTUs for 286 matrices (no threshold)

@@

We investigated whether the available cladistic data for each order was (i) randomly distributed, (ii) overdispersed or (iii) clustered, with respect to phylogeny, using two metrics from community phylogenetics: the Nearest Taxon Index (NTI; \cite{webb2002phylogenies}) and the Net Relatedness Index (NRI; \cite{webb2002phylogenies}). 
NTI is most sensitive to clustering or overdispersion near the tips, whereas NRI is more sensitive to clustering or overdispersion across the whole phylogeny \cite{Cooper2008}. 
Both metrics were calculated using the \texttt{picante} package in R \cite{picante,R}.

NTI \cite{webb2002phylogenies} is based on mean nearest neighbour distance ($MNND$) and is calculated as follows:
  \begin{equation}
    \text{NTI}=-\left(\frac{\overline{MNND}_{obs}-\overline{MNND}_{n}}{\sigma(MNND_{n})}\right)
  \end{equation}
where $\overline{MNND}_{obs}$ is the observed mean distance between each of $n$ taxa with cladistic data and its nearest neighbour with cladistic data in the phylogeny, 
$\overline{MNND}_{n}$ is the mean of 1000 $MNND$ between $n$ randomly drawn taxa, and $\sigma(MNND_{n})$ is the standard deviation of these 1000 random $MNND$ values.
NRI is calculated in the same way, but $MNND$ is replaced by mean phylogenetic distance ($MPD$) as follows:
  \begin{equation}
    \text{NRI}=-\left(\frac{\overline{MPD}_{obs}-\overline{MPD}_{n}}{\sigma(MPD_{n})}\right)
  \end{equation}
where $\overline{MPD}_{obs}$ is the observed mean phylogenetic distance of the tree containing only the $n$ taxa with cladistic data.
Negative NTI and NRI values show that the focal taxa are more overdispersed across the phylogeny than expected by chance, and positive values reflect clustering.

@@

\section{Discussion}
Our results show that although phylogenetic relationships among living mammals are well-resolved (e.g. \cite{BinindaEmonds,meredithimpacts2011}) , most of the data used to build these phylogenies is molecular, and very little cladistic data is available for living mammals compared to fossil mammals (e.g. \cite{O'Leary08022013,ni2013oldest}).
This has implications for building Total Evidence phylogenies containing both living and fossil mammals, as without sufficient cladistic data for living species, fossil placements in these trees are very uncertain \cite{GuillermeCooper}.

The number of living mammalian taxa with no available cladistic data was surprisingly high at the species-level: only six out of 28 orders have a high coverage of taxa with available cladistic data.
This high coverage threshold of 75\% of taxa with available cladistic data represents the minimum amount of data required before missing data has a significant effect on the topology of Total Evidence trees \cite{GuillermeCooper}.
Beyond this threshold, there is considerable displacement of wildcard taxa (\textit{sensu} \cite{kearneyfragmentary2002}) and decreased clade conservation \cite{GuillermeCooper}.
Therefore we expect difficulties in placement of fossil taxa at the species-level in most mammalian orders, but fewer issues at higher taxonomic levels.
This point is important from a practical point of view because of the slight discrepancy between neontological and palaeontological species concepts.
While neontological species are described using morphology, genes, distribution etc.; palaeontological species can be based only on morphological, spatial and temporal data (e.g. \cite{ni2013oldest}).
Therefore, most palaeontological studies use genus as their smallest OTU (e.g. \cite{ni2013oldest,O'Leary08022013}), so data availability at the genus-level in living mammals should be our primary concern when building phylogenies of living and fossil taxa.

@@

%Biology letters various stuff
\section{Ethics statement}
N/A
\section{Data accessibility statement}
All data and analysis code is available on GitHub (\url{https://github.com/TGuillerme/Missing_living_mammals}).
\section{Authors' Contributions}
T.G. and N.C conceived and designed the experiments. T.G. performed the experiments and analysed the data. T.G. and N.C. contributed to the writing of the manuscript. All authors approved the final version of the manuscript.
\section{Competing Interests}
We have no competing interests.
\section{Acknowledgments}
We thank David Bapst, Graeme Lloyd, Nick Matzke and April Wright.
\section{Funding statement}
This work was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal number: 321696).

>>>>

There is an increasing consensus among biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \cite{slaterunifying2013,fritzdiversity2013}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \cite{ronquista2012}, combines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix (e.g. \cite{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014}), producing a phylogeny with living and fossil taxa at the tips. 
A downside of this method is that it requires molecular data for living taxa and morphological data for both living and fossil taxa.
Chunks of this data can be difficult, or impossible, to collect for every taxon in the analysis.
For example, fossils rarely have molecular data and incomplete fossil preservation may restrict the amount of morphological data available.
Additionally, it has become less common to collect morphological characters for living taxa when molecular data is available (e.g. in \cite{slaterphylogenetic2013}, only 13\% of living taxa have coded morphological data).
Unfortunately this missing data can lead to errors in phylogenetic inference.
Simulations show that the ability of the Total Evidence method to recover the correct topology decreases when there is little overlap between morphological data in living and fossil taxa, and that the effect of missing data on topology is greatest when living taxa have few morphological data \cite{GuillermeCooper}.
This is because (1) fossils cannot branch in the correct clade if it contains no morphological data for living taxa; and (2) fossils have a higher probability of branching within clades with more morphological data for living taxa, regardless of whether this is the correct clade \cite{GuillermeCooper}.

@@

%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all cladistic matrices containing any living and/or fossil mammal taxa from three major public databases: MorphoBank (\url{http://www.morphobank.org/} \cite{morphobank}), Graeme Lloyd's website (\url{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\url{https://github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search for matrices that were not uploaded to these databases (see Electronic Supplementary Material (ESM) for a detailed description of the search procedure).
In total, we downloaded 286 matrices containing 5228 unique operational taxonomic units (OTUs). 
We used OTUs rather than species since entries in the matrices ranged from species to families, and standardised the taxonomy as described in the ESM. 
We designated as ``living'' all OTUs that were either present in the phylogeny of \cite{BinindaEmonds} or the taxonomy of \cite{wilson2005mammal}.

Matrices with few characters are problematic when comparing available data among matrices because (1) they have less chance of having overlapping characters with other matrices \cite{wagner2000} and (2) they are more likely to contain specific characters that are not-applicable across large clades (e.g. ``antler ramifications'' is a character that is only applicable to Cervidae \cite{Brazeau2011}).
Therefore we selected only matrices containing \textgreater 100 characters for each OTU.
This threshold was chosen to correspond with the number of characters used in \cite{GuillermeCooper} and \cite{harrisonamong-character2014}.
Results of analyses with no threshold are available in the ESM. 
After removing matrices with \textless 100 characters, we retained 1074 unique living OTUs from 126 matrices. % 1601 unique living OTUs for 286 matrices (no threshold)

@@

We investigated whether the available cladistic data for each order was (i) randomly distributed, (ii) overdispersed or (iii) clustered, with respect to phylogeny, using two metrics from community phylogenetics: the Nearest Taxon Index (NTI; \cite{webb2002phylogenies}) and the Net Relatedness Index (NRI; \cite{webb2002phylogenies}). 
NTI is most sensitive to clustering or overdispersion near the tips, whereas NRI is more sensitive to it across the whole phylogeny \cite{Cooper2008}. 
Both metrics were calculated using the \texttt{picante} package in R \cite{picante,R}.

NTI \cite{webb2002phylogenies} is based on mean nearest neighbour distance ($MNND$) and is calculated as follows:
  \begin{equation}
    \text{NTI}=-\left(\frac{\overline{MNND}_{obs}-\overline{MNND}_{n}}{\sigma(MNND_{n})}\right)
  \end{equation}
where $\overline{MNND}_{obs}$ is the observed mean distance between each of $n$ taxa with cladistic data and its nearest neighbour with cladistic data in the phylogeny, 
$\overline{MNND}_{n}$ is the mean of 1000 $MNND$ between $n$ randomly drawn taxa, and $\sigma(MNND_{n})$ is the standard deviation of these 1000 random $MNND$ values.
NRI is calculated in the same way, but using the mean phylogenetic distance ($MPD$):
  \begin{equation}
    \text{NRI}=-\left(\frac{\overline{MPD}_{obs}-\overline{MPD}_{n}}{\sigma(MPD_{n})}\right)
  \end{equation}
where $\overline{MPD}_{obs}$ is the observed mean phylogenetic distance of the tree containing only the $n$ taxa with cladistic data.
Negative NTI and NRI values show that the focal taxa are more overdispersed across the phylogeny than expected by chance, and positive values reflect clustering.

@@

\section{Discussion}
Our results show that although phylogenetic relationships among living mammals are well-resolved (e.g. \cite{BinindaEmonds,meredithimpacts2011}), most of the data used to build these phylogenies is molecular, and few cladistic data are available for living mammals compared to fossils (e.g. \cite{O'Leary08022013,ni2013oldest}).
This has implications for building Total Evidence phylogenies containing both living and fossil mammals, as without sufficient cladistic data for living species, fossil placements in these trees are very uncertain \cite{GuillermeCooper}.

The number of living mammalian taxa with no available cladistic data was surprisingly high at the species-level: only six out of 28 orders have a high coverage of taxa with available cladistic data.
This high coverage threshold of 75\% of taxa with available cladistic data represents the minimum amount of data required before missing data has a significant effect on the topology of Total Evidence trees \cite{GuillermeCooper}.
Beyond this threshold, there is considerable displacement of wildcard taxa and decreased clade conservation \cite{GuillermeCooper}.
Therefore we expect difficulties in placement of fossils at the species-level in most mammalian orders, but fewer issues at higher taxonomic levels.
This point is important from a practical point of view because of the slight discrepancy between neontological and palaeontological species concepts.
While neontological species are described using morphology, genes, distribution etc.; palaeontological species can be based only on morphological, spatial and temporal data (e.g. \cite{ni2013oldest}).
Therefore, most palaeontological studies use the genus as their smallest OTU (e.g. \cite{ni2013oldest,O'Leary08022013}), so data availability at the genus-level in living mammals should be our primary concern when building phylogenies of living and fossil taxa.

@@

%Biology letters various stuff
\section{Ethics statement}
N/A
\section{Data accessibility statement}
All data and code are available on GitHub (\url{https://github.com/TGuillerme/Missing_living_mammals}).
\section{Authors' Contributions}
T.G. and N.C designed the study. T.G. analysed the data. T.G. and N.C. wrote the the manuscript.
\section{Competing Interests}
We have no competing interests.
\section{Acknowledgments}
We thank David Bapst, Graeme Lloyd, Nick Matzke and April Wright.
\section{Funding statement}
This work was funded by a European Commission CORDIS Seventh Framework Programme (FP7) Marie Curie CIG grant (proposal number: 321696).

\section{References}
